---
title: "Abalone Age (CRISP-DM)"
author: "David"
date: May 2021
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: yeti
  pdf_document: default
---

# Problem statement
Dataset source: https://archive.ics.uci.edu/ml/datasets/Abalone


Create a tool which can be used by researchers in the field to estimate the age of an abalone species based on measurements. This problem my be treated as regression/classification - I will treat it as regression.

```{r}



```

# Data understanding and data preparation
```{r, message=F, warning=F}

library(dplyr)
library(data.table)
library(caret)
library(mlbench)
library(Metrics)
library(ggfortify)
library(Cubist)

```

# Modeling and evaluation
## Exploring feature importance
```{r}



```

## Feature selecton (caret)
```{r}



```

### Multiple linear regression
```{r}



```

### Model tree
```{r}



```

### Principal component analysis
```{r}



```

#### Model summary
While the model tree model provided superior results compared to linear regression, fitting a model using the first 3 principal components provided excellent test set accuracy. The PCA model should be used in production.

# Deployment
```{r}



```

